---
title: "Feb13_pres"
format: revealjs
editor: visual
---

## A summary of Findings + Ideas 

Author : Ethan Heck

## Survival Model: Using Mistry's work as a starting point
```{r, echo=FALSE}

# mac Syntax
# load("~/Desktop/Survival Intern Project/Survival/atlas/dat/atlasDataClean.RData")

#windows syntax
load("~/atlas/dat/atlasDataClean.RData")

#package 
library(survival)
library(survminer) 
library(ggplot2) 
library(dplyr)
```

```{r, echo=TRUE}

####Getting complete cases for survival######
atlasDataClean <- subset(atlasDataClean,atlasDataClean$survivalMonths != "NA")

#surv model
fit <- survfit(Surv(time = as.numeric(survivalMonths), event = mortality) ~ grade , 
               data = atlasDataClean)

ggsurvplot(
  fit,
  data = atlasDataClean,
  risk.table = TRUE)
```

## Two ways of plotting KP curves (so far)


```{r, echo=TRUE}
# Example survival model (stratified by 4 groups)
fit <- survfit(Surv(time = as.numeric(survivalMonths), event = mortality) ~ grade , 
               data = atlasDataClean) 

# Convert survfit object to tidy data
surv_df <- as.data.frame(summary(fit)[c("time", "surv", "strata", "lower", "upper")])

# Rename columns for convenience
colnames(surv_df) <- c("time", "survival", "group", "lower", "upper")



# Create the Kaplan-Meier plot using ggplot
ggplot(surv_df, aes(x = time, y = survival, color = group, fill = group)) +
  geom_step(size = 1.2) +  # Kaplan-Meier survival curve
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2, color = NA) +  # Confidence intervals
  labs(
    title = "Survival Probability (Months) ",
    x = "Time (Months)",
    y = "Survival Probability",
    color = "Group",
    fill = "Group"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(hjust = 0.5, size = 16, face = "bold"),
    legend.position = "right"
  )

#same plot using ggsurvfit
ggsurvplot(
  fit,
  data = atlasDataClean,
  conf.int = TRUE,
  pval = TRUE,
  xlab = "Time (Months)",
  ylab = 'Survival Probability',
  risk.table = TRUE)
```

---
## Two ways of plotting : Continued
-Using ggplot
    -Have to convert surv obj to df
    -can switch names if preferred
    -then plot with ggplot directly
    -can work w ggplot customizations directly
-Using ggsurvplot
    -takes surv object directly
    -can use most ggoptions within the plot directly
    -preferred method*
    

## Outputting Median Survival Times
```{r, echo=TRUE}
##for whole survival table 
summary(fit)$table


## for just median survival times 
summary(fit)$table[, "median"]

```

-Note: group 1 has "NA" for median survival time... why?
    - this is because in this model, where survival is stratified by grade,
    the survival probability of those with a grade 1 tumor never falls at or below 0.5.
    -By definition, median survival time is the time at which the probability of survival is 0.5. 
    
## Hazard Ratios - CoxPH models

```{r, echo=TRUE}
#cox model --- uses same surv syntax as fit object above
cox_model <- coxph(Surv(time = as.numeric(survivalMonths), event = mortality) ~ grade, 
                   data = atlasDataClean)
summary(cox_model)

#Have to exponentiate model coefficients (betas) to get Hazard ratios 
hr_values <- exp(coef(cox_model))


print(hr_values)

```

-First question? - why does grade 1 have no value
    -this is because grade 1 is the reference group for creating hazard ratios
    -this can be changed if preferred using "relevel"
- Interpretting hazard ratios?
    -Grade 2 [5.59] - those with grade 2 tumors are 5.6x likely to experience an event (death) than grade 1
    -Grade 3 [8.14] - those with a grade 3 tumor are 8.14x likely to experience an event (death) than
grade 1
    -Grade 4 [19.99] - those with a grade 4 tumor are 20x likely to experience an event (death) than 
grade 1
