

# survival analysis
```{r}

freq(is.na(atlasDataClean$survivalMonths))

myView(subset(atlasDataClean, !is.na(survivalMonths)))


myView(subset(atlasData, !is.na("Overall survival (months)")))


sink('togrok'); freq(atlasData$`Overall survival (months)`); sink()

# Filter Overall survival (months) to only include values with special characters 
# (excluding decimal points)
special_char_values <- atlasData[grepl("[^a-zA-Z0-9 \\.]", atlasData$`Overall survival (months)`), "Overall survival (months)"]

# Display the filtered values
special_char_values

# Filter rows where Overall survival (months) contains special characters (excluding decimal points)
special_char_rows <- atlasData[grepl("[^a-zA-Z0-9 \\.]", atlasData$`Overall survival (months)`), ]

# Display the filtered rows
special_char_rows

special_char_values <- atlasData[grepl("[-<>]", atlasData$`Overall survival (months)`), "Overall survival (months)"]
special_char_values

special_char_values <- atlasData[grepl("[^0-9]", atlasData$`Overall survival (months)`), "Overall survival (months)"]
special_char_values


names(atlasDataClean)
subgroup, country

```


```{r}
atlasDataCleanSurvival <- atlasDataClean %>%
  filter(!is.na(survivalMonths))


```


```{r}
# Function to count occurrences for each variable, one at a time
count_by_variables <- function(data, group_vars) {
  results_list <- list()
  
  for (var in group_vars) {
    cat("Counting by:", var, "\n") # Optional: print progress
    
    results_list[[var]] <- data %>%
      filter(!is.na(survivalMonths)) %>%
      group_by(across(all_of(var))) %>%
      count(sort = TRUE)
  }
  
  return(results_list)
}

# Define your list of variables to group by
# group_vars <- c("cancer_type", "gender", "age_group", "stage", "treatment_type", 
#                 "smoking_status", "tumor_site", "histology") # Add all 40 variables

exclude_vars <- c("filename", "dataID", "sampleID", 
                  "classifierLightGBM", "classifierRandomForest", 
                  "classifierEuclidean", "classifierTSNE",
                  "tsne1", "tsne2", "age", "survivalMonths")

# group_vars <- names(atlasDataClean)

group_vars <- setdiff(names(atlasDataClean), exclude_vars)

# Run the function
count_results <- count_by_variables(atlasDataClean, group_vars)

# To view results for a specific variable:
count_results[["cancer_type"]]

sink("variable_counts.txt")

# To print all results:
for (var in group_vars) {
  cat("\n\n===", var, "===\n")
  print(count_results[[var]])
}

sink()
# Alternatively, if you want to apply a function to each variable and return the results directly:
# map_results <- group_vars %>%
#   purrr::map(~{
#     atlasDataClean %>%
#       filter(!is.na(survivalMonths)) %>%
#       group_by(across(all_of(.x))) %>%
#       count(sort = TRUE)
#   })

# # Name the results with their respective variable names
# names(map_results) <- group_vars
```


```{r}
freq(atlasDataClean$diagnosisClass)

unique(atlasDataClean$sampleID) %>% length()

group_by(atlasDataClean, sampleID) %>% count(sort = T) %>% filter(n > 1)
```

# assign missing diagnosis class to testing data set?

```{r, groupings}

atlasDataClean %>% 
  group_by(diagnosisClass, diagnosisFinal) %>%
  summarise(n = n()) %>% 
  arrange(diagnosisFinal)


```

# attempt to use the diagnosisFinal to assign missing diagnosisClass
```{r}
# Create a reference mapping from existing valid entries
diagnosis_mapping <- atlasDataClean %>%
  filter(diagnosisClass != "#N/A" & !is.na(diagnosisClass)) %>%
  distinct(diagnosisFinal, diagnosisClass)

# Apply the mapping to fill in #N/A values
atlasDataClean2 <- atlasDataClean %>%
  # left join safer; but for now, same output as inner_join
  left_join(diagnosis_mapping, by = "diagnosisFinal", suffix = c("", "_new")) %>% 
  mutate(diagnosisClass = ifelse(diagnosisClass == "#N/A" | is.na(diagnosisClass),
                                diagnosisClass_new,
                                diagnosisClass)) %>%
  select(-diagnosisClass_new)

rm(diagnosis_mapping)
```

# understand
# gene ontology biological processes (BP) cannot be merged into a DF with 20360 rows because "one to many" is the relationship of ENTREZID and BP

```{r}
# jan 21 2025
# understand gene annotations

tellMe <- names(mapped_genes) %>% 
  unique()

sample(tellMe, 50)


glimpse(atlasDataClean)
glimpse(geneExpressionData[1:25, 1:5])
glimpse(gene_annotations)
rownames(geneExpressionData[1:25, 1:5])

# jan 23 2025

# Load necessary library
library(purrr)

# Assuming 'mapped_genes' is a list or a vector where each element is like mapped_genes[1]
result <- map(mapped_genes, function(x) {
  if(is.list(x) && length(x) == 1) {
    # If x is a list with one element, apply the operation
    return(length(unlist(x[[1]])))
  } else {
    # If x does not match expected structure, return NA or another fallback value
    return(NA)
  }
})

# Convert the result to integers, assuming all results are numeric or can be converted
integer_values <- as.integer(unlist(result))

# Check if we have 6394 elements
if(length(integer_values) == 6394) {
  print("Successfully processed 6394 elements.")
} else {
  warning("The number of processed elements does not match 6394.")
}

# Print or use integer_values as needed
print(integer_values)

# Calculate the sum of the lengths of all elements in mapped_genes
total_elements <- sum(lengths(mapped_genes))

print(total_elements)

# n = 406318.. ENTREZIDs in multiple categories?
# Combine all elements into one vector, then find unique values and count them
unique_elements <- unique(unlist(mapped_genes))
total_unique_elements <- length(unique_elements)

print(total_unique_elements)

# Combine all elements into one vector
all_values <- unlist(mapped_genes)

# If you want to see the first few elements of this vector for verification:
print(head(all_values))
li(all_values, rownames(geneExpressionData))

# check gene annotations

library(rentrez)

# Function to check if an Entrez ID is valid
check_entrez_id <- function(id) {
  tryCatch({
    # Query NCBI for the gene ID
    result <- entrez_summary(db="gene", id=id)
    return(TRUE)
  }, error = function(e) {
    # If an error occurs (e.g., ID not found), return FALSE
    return(FALSE)
  })
}

# Example IDs, replace with your full list
ids_to_check <- filter(gene_annotations, is.na(SYMBOL)) %>% pull(ENTREZID)

# Check validity
valid_ids <- sapply(ids_to_check, check_entrez_id)
print(valid_ids)

filter(gene_annotations, is.na(SYMBOL)) %>% glimpse()

#####

```

# rentrez

```{r}
library(rentrez)

fetch_gene_info <- function(id) {
  result <- entrez_summary(db="gene", id=id)
  return(data.frame(ENTREZID = id, 
                    SYMBOL = result$name, 
                    GENENAME = result$description))
}

# Fetch info for missing IDs
missing_ids <- gene_annotations$ENTREZID[is.na(gene_annotations$SYMBOL)]
new_annotations <- do.call(rbind, lapply(missing_ids, fetch_gene_info))

# Merge with existing annotations
# gene_annotations <- rbind(gene_annotations[!is.na(gene_annotations$SYMBOL), ], new_annotations)


##

# library(AnnotationHub)
# Warning in install.packages :
#   package ‘AnnotationHub’ is not available for this version of R
# ah <- AnnotationHub()
# # Query for human gene databases
# query(ah, "Homo sapiens")  # Look for relevant databases
```

# rentrez II

```{r}
## RUN TIME: > 45 minutes

# library(rentrez)
# 
# fetch_gene_info <- function(entrez_ids) {
#   # Initialize an empty list to store results
#   results <- list()
# 
#   # Loop through each Entrez ID
#   for (id in entrez_ids) {
#     try({
#       # Fetch gene summary for each ID
#       result <- entrez_summary(db="gene", id=id)
#       results[[id]] <- data.frame(ENTREZID = id, 
#                                   SYMBOL = result$name, 
#                                   GENENAME = result$description)
#     }, silent = TRUE)  # Silent error handling to continue with next ID if one fails
#   }
# 
#   # Combine all results into one data frame
#   if (length(results) > 0) {
#     do.call(rbind, results)
#   } else {
#     data.frame(ENTREZID = character(), SYMBOL = character(), GENENAME = character())
#   }
# }
# 
# # Use the function with the rownames of geneExpressionData
# gene_info <- fetch_gene_info(rownames(geneExpressionData))
```

# validated appender

```{r}
# Assuming 'mapped_genes' is a list where keys are names and values are vectors of ENTREZIDs

# Convert the list to a data frame where each ENTREZID can have multiple names
library(data.table)

# Create a data frame from the list
mapped_genes_df <- data.table::rbindlist(
  lapply(names(mapped_genes), function(name) {
    data.frame(ENTREZID = mapped_genes[[name]], NAME = name)
  }), fill = TRUE
)

# Convert ENTREZID to character if it's not already (to match with gene_annotations)
mapped_genes_df$ENTREZID <- as.character(mapped_genes_df$ENTREZID)

# Perform a left join (or merge) to add all NAMEs from mapped_genes_df to gene_annotations
# where ENTREZID matches. This will create a row for each name per ENTREZID.
result <- merge(gene_annotations, mapped_genes_df, by = "ENTREZID", all.x = TRUE)

# If you want to aggregate names for each ENTREZID (since one ENTREZID can match multiple names)
result <- result %>%
  dplyr::group_by(ENTREZID, SYMBOL, GENENAME) %>%
  dplyr::summarise(NAME = paste(unique(NAME), collapse = ", "), .groups = 'drop')

# Display the head of the result
head(result)
```

## this would be next step (see note next to 'understand')

```{r}
library(dplyr)

# Update result SYMBOLs where they are NA and there's a matching ENTREZID in new_annotations
result <- result %>%
  left_join(new_annotations %>% select(ENTREZID, SYMBOL), by = "ENTREZID") %>%
  mutate(SYMBOL = if_else(is.na(SYMBOL.x), SYMBOL.y, SYMBOL.x)) %>%
  select(-SYMBOL.x, -SYMBOL.y)

# Verify changes
sum(is.na(result$SYMBOL))  # Should be less than 96 if matches exist

# gene groupings

# Summary
# To find gene groupings:

# Use predefined databases (MSigDB, GO, KEGG, Reactome).
# Analyze your data for statistical groupings (e.g., variability, DEGs).
# Map genes to functional categories with Bioconductor tools.
# Create custom groupings based on literature or expert input.
# Organize these groupings into a structure suitable for Shiny selectors (lists or data frames).

# org.Hs.eg.db

```


```{r}
# BiocManager::install("org.Hs.eg.db")

library(org.Hs.eg.db)

# Column Availability
# cols_available <- columns(org.Hs.eg.db)

# Map Entrez IDs to Gene Symbols and Names
gene_annotations <- AnnotationDbi::select(org.Hs.eg.db,
                                          keys = rownames(geneExpressionData),
                                          columns = c("SYMBOL", "GENENAME"),
                                          # columns = cols_available,
                                          keytype = "ENTREZID")

# head(gene_annotations)
# select(gene_annotations, ONTOLOGY) %>% pull %>% unique()
# freq(gene_annotations$ONTOLOGY)
# table(gene_annotations$ONTOLOGY, useNA = "ifany")
# handle unmapped IDs (none! found)
# unmapped_ids <- setdiff(rownames(geneExpressionData), gene_annotations$ENTREZID)
# print(unmapped_ids)


# Group genes

# biological pathways
# BiocManager::install("clusterProfiler")
library(clusterProfiler)

# Map genes to KEGG pathways (400 error)
# kegg_annotations <- enrichKEGG(gene = rownames(geneExpressionData),
#                                 organism = "hsa", # homo sapiens
#                                 keyType = "ENTREZID")
# head(kegg_annotations)

# gene ontology terms
go_annotations <- enrichGO(gene = rownames(geneExpressionData),
                            OrgDb = org.Hs.eg.db,
                            ont = "BP", # Biological Process (or: “BP”, “MF”, “CC”, “ALL”)
                            keyType = "ENTREZID")
head(go_annotations)

# Extract gene-to-GO mapping
# This creates a named list where:
# 
# The names are GO terms (descriptions).
# The values are comma-separated gene IDs.
go_to_genes <- setNames(go_annotations@result$geneID, go_annotations@result$Description)

# Split the gene IDs into vectors
go_to_genes_list <- lapply(go_to_genes, function(x) unlist(strsplit(x, "/")))

# Find overlap with your gene expression data
mapped_genes <- lapply(go_to_genes_list, function(genes) {
  intersect(genes, rownames(geneExpressionData))
})

# Summarize expression by GO terms
go_expression <- sapply(names(mapped_genes), function(go_term) {
  genes <- mapped_genes[[go_term]]
  if (length(genes) > 0) {
    rowMeans(geneExpressionData[genes, , drop = FALSE])  # Mean expression for the term
  } else {
    NA  # No mapped genes for this GO term
  }
})

save(go_expression, gene_annotations, go_annotations, go_to_genes_list, go_to_genes, mapped_genes, file = 'dat/annotations.RData')

```

# check mapped_genes

```{r}
# Assuming mapped_genes is a list where each element can have different lengths
lengths <- numeric(6394)  # Pre-allocate space for efficiency
for (i in 1:6394) {
  # Check if the list index exists to avoid errors
  if (i <= length(mapped_genes)) {
    lengths[i] <- length(mapped_genes[[i]])
  } else {
    lengths[i] <- NA  # or some default value if the index is out of bounds
  }
}
# lengths now contains the length of each element in mapped_genes
# No print statement needed; the vector 'lengths' is ready for use or further operations


## check uniqueness
# Combine all elements into one list
all_elements <- unlist(mapped_genes, recursive = FALSE)

# Find the distinct count of all items
distinct_count <- length(unique(all_elements))

# Optional: If you want to see the unique values
# unique_values <- unique(all_elements)

# Print or use the distinct count as needed
print(distinct_count)

#  > 14000/20360

#  [1] 0.6876228 (of all genes represented in mapped_genes)
```

## pheatmap

```{r}
library(pheatmap)

# Remove GO terms with no mapped genes
# go_expression_clean <- go_expression[!is.na(go_expression), ] # threw error
go_expression_clean <- go_expression[!is.na(go_expression)]
str(go_expression_clean) # list. needs to be numeric matrix or df
# Convert the list to a numeric matrix
go_expression_clean <- do.call(rbind, lapply(go_expression_clean, as.numeric))

# Check the resulting structure
str(go_expression_clean)


# Plot heatmap
pheatmap(go_expression_clean, cluster_rows = TRUE, cluster_cols = TRUE,
         main = "GO Term-Level Expression")


library(heatmaply) # way too slow; dont run again

heatmaply(go_expression_clean, 
          k_row = 10,  # Cluster rows into 10 groups
          k_col = 10,  # Cluster columns into 10 groups
          main = "GO Term-Level Expression")

library(ggplot2)

# Summarize mean expression for GO terms
mean_expression <- rowMeans(go_expression_clean)

# Barplot for top 20 GO terms
top_mean_expression <- sort(mean_expression, decreasing = TRUE)[1:20]
ggplot(data = data.frame(GO = names(top_mean_expression), Expression = top_mean_expression),
       aes(x = reorder(GO, Expression), y = Expression)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(title = "Top 20 GO Terms by Mean Expression", x = "GO Term", y = "Mean Expression")

```

## Ensembl

```{r}
library(org.Hs.eg.db)
library(clusterProfiler)
library(biomaRt)
library(ggplot2)
library(pheatmap)

# Assuming 'geneExpressionData' is your data frame or matrix with Entrez IDs as rownames

# Fetch gene symbols from Ensembl
ensembl = useMart("ensembl", dataset="hsapiens_gene_ensembl")
geneSymbols <- getBM(attributes = c("entrezgene_id", "hgnc_symbol"),
                     filters = "entrezgene_id", 
                     values = rownames(geneExpressionData),
                     # values = test_ids,
                     mart = ensembl)

# output = 18906 / 20360 = 0.93

# Create unique gene symbols to avoid duplicate row name issues
# Here, we ensure uniqueness by considering both Entrez ID and the original symbol
unique_symbols <- paste(geneSymbols$hgnc_symbol, geneSymbols$entrezgene_id, sep = "_")
geneSymbols$unique_symbol <- make.unique(unique_symbols, sep = "_")

# Set row names with these unique symbols
rownames(geneExpressionData) <- setNames(geneSymbols$unique_symbol, 
                                         geneSymbols$entrezgene_id)[rownames(geneExpressionData)]

# Use original Entrez IDs for GO and KEGG analysis
entrez_ids <- rownames(geneExpressionData[1:100,])

# Perform GO enrichment
go_enrich <- enrichGO(gene = entrez_ids,
                      OrgDb = org.Hs.eg.db,
                      ont = "BP",
                      pAdjustMethod = "BH",
                      pvalueCutoff = 0.05,
                      qvalueCutoff = 0.2,
                      readable = TRUE)

# Display GO enrichment results
print(head(go_enrich, n = 10))

# KEGG pathway analysis
kegg_enrich <- enrichKEGG(gene = entrez_ids,
                          organism = 'hsa',  
                          pvalueCutoff = 0.05)

# Display KEGG results
print(head(kegg_enrich))

# Custom grouping for ZNF genes
znf_genes <- grep("^ZNF", rownames(geneExpressionData), value = TRUE)
if(length(znf_genes) > 0) {
  znf_group <- data.frame(Gene = znf_genes, Group = "ZNF")
  print(znf_group)

  # Visualization of GO enrichment
  print(dotplot(go_enrich, showCategory = 30) + ggtitle("GO Enrichment Analysis"))

  # Heatmap for ZNF genes
  pheatmap(geneExpressionData[znf_genes,], 
           show_rownames = TRUE, 
           main = "Expression of ZNF Genes",
           fontsize_row = 6)
} else {
  print("No ZNF genes found or they might have unique suffixes.")
}
```


```{r}
# cross reference with known dB
# Select a small subset of IDs
test_ids <- head(rownames(geneExpressionData))

# Map these IDs to gene symbols
mapped_genes <- AnnotationDbi::select(org.Hs.eg.db, 
                                      keys = test_ids,
                                      # keys = rownames(geneExpressionData),
                                      columns = c("SYMBOL", "GENENAME", "GO", "PATH"), 
                                      keytype = "ENTREZID")
print(mapped_genes)

# map genes to GO biological processes

# enrichGO(gene = test_ids, OrgDb = org.Hs.eg.db, ont = "BP")


# merge
# Example of merging annotations with gene expression data
# merged_data <- merge(geneExpressionData, annotations, 
#                      by.x = "rownames", by.y = "ENTREZID", all.x = TRUE)

```

## kegg

```{r}
library(KEGGREST)
library(clusterProfiler)

# Convert Entrez IDs to KEGG IDs
entrez_ids <- rownames(geneExpressionData)
# kegg_ids <- keggConv("kegg", "hsa")  # Converts human Entrez IDs to KEGG IDs
kegg_ids <- keggConv("hsa", "ncbi-geneid")

# Filter your Entrez IDs to match KEGG
valid_kegg_ids <- intersect(entrez_ids, gsub(".*:", "", names(kegg_ids)))

# Run enrichment analysis with valid KEGG IDs
kegg_annotations <- enrichKEGG(gene = valid_kegg_ids,
                                organism = "hsa", 
                                keyType = "kegg")

# length of intersection between keggID and ENTREZID in this dataset = 16949 / 20360 = 0.83

```

# examine

## non-tumor?

```{r cross-reference and explore}

examine <- geneExpressionData %>% 
  select(AGZ_001_U133_2.CEL)

examineOther <- atlasData %>% 
  filter(Filename %in% 'AGZ_001_U133_2.CEL')

rownames(examine) %>% length() %>% unique()

n_distinct(atlasData$Filename) # matches 7375

atlasData %>% 
  group_by(
    Filename
  ) %>% 
  count()

# List of non-tumor categories from the image
non_tumor_categories <- toupper(c("Cerebellum", "Cerebellum, Fetal", "Choroid Plexus", 
                          "CNS", "Nerve Ganglia", "Peri. Nerve", "Pituitary", 
                          "Retina", "Retina, Fetal"))


# Create the binary indicator variable
atlasData2$is_tumor <- ifelse(
  # atlasData2$`Final diagnosis` %in% non_tumor_categories, 
  atlasData2$`Diagnosis class` %in% 'Non-tumor',
  0, 
  1
)

# Check the frequency to ensure the counts match
freq(atlasData2$is_tumor)

sample(atlasData$'Sample ID', 250) %>% freq

atlasDataClean %>% 
  mutate(
    x = str_sub(filename, 1, 3),
    .keep = 'used'
  ) %>% 
  group_by(x) %>% 
  count(sort = T)


```



```{r}

glimpse(geneExpressionData)
glimpse(atlasData)

```



## freqs

```{r freqs}


map(atlasData[, "Diagnosis class", drop = F], ~ freq(.x))
map(atlasData[, "Diagnosis", drop = F], ~ freq(.x))
map(atlasData[, "Final diagnosis", drop = F], ~ freq(.x))
map(atlasData[, "Age (grouped)", drop = F], ~ freq(.x))

map(select(atlasData, contains('classifier')), drop = F, ~ freq(.x))

map(atlasData[, , drop = F], ~ freq(.x) %>% head())
map(atlasDataClean[, , drop = F], ~ freq(.x) %>% head())
sink('dat/all-freqs'); map(atlasData, ~ freq(.x)); sink()
sink('dat/all-freqs2'); map(atlasDataClean, ~ freq(.x)); sink()
sink('out/all-freqs3'); map(atlasDataClean, ~ freq(.x)); sink()
sink('out/all-freqs4'); map(select(atlasDataClean, -c('filename', 'dataID', 'sampleID', 'tsne1', 'tsne2')), ~ freq(.x)); sink()

atlasData %>% filter(as.logical(is_tumor)) %>% pull('Final diagnosis') %>% freq
atlasData %>% filter(as.logical(is_tumor)) %>% pull('Diagnosis class') %>% freq

atlasDataClean %>% 
  pull(diagnosisFinal) %>% 
  freq()
atlasDataClean %>% 
  pull(diagnosis) %>% 
  freq()
atlasDataClean %>% 
  pull(diagnosisClass) %>% 
  freq()


```




# extra

## read

```{r}
# ubuntu
# geneExpressionData <- readRDS('/mnt/KDPH/PQI/Division Office/PQI Data Analytics/BrainTumorAtlas-main/Gene Expression Data/ExpData.rds')
# atlasData <- read_csv('/mnt/KDPH/PQI/Division Office/PQI Data Analytics/BrainTumorAtlas-main/Atlas_Data.csv')

# windows

# geneExpressionData <- readRDS("P:/PQI/Division Office/PQI Data Analytics/BrainTumorAtlas-main/Gene Expression Data/ExpData.rds")
# # atlasData <- read_csv('dat/Atlas_Data.csv')
# save(atlasData, file = 'dat/atlasData.RData')
```

## vec

```{r}

# Define the vector for both ageGroup levels and labels
# age_groups <- c("FETAL", "0-5YRS", "5-10YRS", "10-20YRS", "20-40YRS", "40-60YRS", "60-80YRS", "80+YRS")


```


## NA-check

```{r na-check}
library(tidyverse)

# Determine which columns are of double type
double_columns <- geneExpressionData %>%
  summarise(across(everything(), is.double)) %>%
  pivot_longer(everything(), names_to = "column", values_to = "is_double") %>%
  filter(is_double) %>%
  pull(column)

# Check for NA values in these double columns
na_check <- geneExpressionData %>%
  select(all_of(double_columns)) %>%
  summarise(across(everything(), ~any(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "column", values_to = "has_na") %>%
  filter(has_na)

# Display the result
if (nrow(na_check) > 0) {
  print("Columns with NA values in double type:")
  print(na_check$column)
} else {
  print("No NA values found in any double type columns.")
}



```

# henry-adapted


```{r adapt-from-hn}
# #atlas<-readRDS("C:/Users/henry/Downloads/ExpData.rds")
# mean.frame <- data.frame(colMeans(geneExpressionData))
# jde.mean<- data.frame(mean.frame[1:170,])
# # jde.hist <- read.csv("C:/Users/henry/Downloads/hist.csv")
# hist <- read_csv("hist.csv")
# jde.data<- bind_cols(jde.mean,hist)
# colnames(jde.data)[colnames(jde.data) == "mean.frame.1.170..."] <- "Mean"
# boxplot(Mean ~ Histology, data = jde.data,
#         main = "Boxplot by Histology", 
#         xlab = "Histology", 
#         ylab = "mRNA Expression (log2)")
```



```{r adapt-from-hn}


# # jde.surv<- read.csv("C:/Users/henry/Downloads/jde survival.csv")
# jde_survival <- read_csv("jdesurv.csv")
# jde.survival$Age.Groups <- jde_survival$`Age (grouped)`
# 
# jde.survival<- jde_survival[, c("Vital status","Overall survival (months)","Original histology"), drop=F]
# # jde.survival$`Overall survival (months)` <- as.numeric(jde.survival$`Overall survival (months)`)
# fit<- survfit(Surv(`Overall survival (months)`,`Vital status`)~strata(Age.Groups), data=jde.survival)
# plot(fit, xlab="Time", ylab="Survival", main="JDE Data", col=c(1:4), lty=c(1),conf.int = F)
# legend("bottomleft",title = "Age Groups", 0.7, legend=c("10-20","20-40","40-60","60-80"), lty=1, col=1:4)


```





# fig1

```{r}
library(ggplot2)
library(dplyr)

# First, let's summarize the data to get counts for each diagnosis
diagnosis_counts <- atlasData %>%
  count(`Final diagnosis`, sort = TRUE) %>%
  filter(n > 50)  # Only show labels for diagnoses with more than 50 occurrences

# Now, compute the position for labels to avoid overlap
diagnosis_positions <- atlasData %>%
  group_by(`Final diagnosis`) %>%
  summarise(
    tsne1 = median(tsne1),
    tsne2 = median(tsne2)
  ) %>%
  inner_join(diagnosis_counts, by = "Final diagnosis")

# Create the plot
ggplot(atlasData, aes(x = tsne1, y = tsne2)) +
  geom_point(aes(color = `Final diagnosis`), alpha = 0.7, size = 2) +
  geom_text(data = diagnosis_positions, 
            aes(label = `Final diagnosis`, x = tsne1, y = tsne2),
            size = 3, check_overlap = TRUE, vjust = 1, hjust = 1) +  # Adjust size and position
  theme_minimal() +
  labs(title = "t-SNE Visualization of Tumor Types with Diagnosis Labels",
       x = "t-SNE 1",
       y = "t-SNE 2",
       color = "Diagnosis") +
  theme(legend.position = "none")  # Remove legend since we're using labels

# Save the plot if you want
# ggsave("tsne_plot_with_labels.png", width = 15, height = 10, dpi = 300)
```

```{r}
library(ggplot2)
library(dplyr)
library(ggrepel)  # For better label placement

# Compute the position for labels
diagnosis_positions <- atlasData %>%
  group_by(`Final diagnosis`) %>%
  summarise(
    tsne1 = median(tsne1),
    tsne2 = median(tsne2)
  )

# Create the plot with all labels
ggplot(atlasData, aes(x = tsne1, y = tsne2)) +
  geom_point(aes(color = `Final diagnosis`), alpha = 0.7, size = 2) +
  geom_text_repel(data = diagnosis_positions, 
                  aes(label = `Final diagnosis`, x = tsne1, y = tsne2),
                  size = 3, 
                  max.overlaps = Inf,  # Show all labels
                  box.padding = 0.5, 
                  point.padding = 0.5, 
                  segment.color = 'grey50') +
  theme_minimal() +
  labs(title = "t-SNE Visualization of Tumor Types with All Diagnosis Labels",
       x = "t-SNE 1",
       y = "t-SNE 2") +
  theme(legend.position = "none")  # Remove legend since we're using labels

# Save the plot if you want
# ggsave("tsne_plot_all_labels.png", width = 15, height = 10, dpi = 300)

```

```{r}
# blue/red no/yes tumor?

# Calculate median positions for labels
diagnosis_positions <- atlasData2 %>%
  group_by(`Final diagnosis`) %>%
  summarise(
    tsne1 = median(tsne1),
    tsne2 = median(tsne2)
  )

# Create the plot with all labels and color by grok_final
ggplot(atlasData2, aes(x = tsne1, y = tsne2)) +
  geom_jitter(aes(color = factor(is_tumor)), alpha = 0.7, size = 2) +
  geom_text_repel(data = diagnosis_positions, 
                  aes(label = `Final diagnosis`, x = tsne1, y = tsne2),
                  size = 3, 
                  max.overlaps = Inf,  # Show all labels
                  box.padding = 0.5, 
                  point.padding = 0.5, 
                  segment.color = 'grey50') +
  scale_color_manual(values = c("0" = "blue", "1" = "red")) +
  theme_minimal() +
  labs(title = "t-SNE Visualization of Tumor Types with All Diagnosis Labels",
       x = "t-SNE 1",
       y = "t-SNE 2",
       color = "Grok Final") +
  theme(legend.position = "none")

```

# euclidean
```{r}
# Load required libraries
library(dplyr)
library(ggplot2)
library(ggrepel)

# Assuming atlasData contains columns: tsne1, tsne2, `Euclidean classifier`, grok_final

# Median imputation for NA values in `Euclidean classifier`
atlasData$`Euclidean classifier imputed` <- ifelse(is.na(atlasData$`Euclidean classifier`), 
                                                   median(atlasData$`Euclidean classifier`, na.rm = TRUE), 
                                                   atlasData$`Euclidean classifier`)

# Calculate median positions for each diagnosis for labeling
diagnosis_positions <- atlasData %>%
  group_by(`Final diagnosis`) %>%
  summarise(
    tsne1 = median(tsne1),
    tsne2 = median(tsne2)
  )

# Create the 2D plot with Euclidean classifier
ggplot(atlasData, aes(x = tsne1, y = tsne2)) +
  geom_point(aes(color = factor(grok_final)), alpha = 0.7, size = 2) +
  geom_text_repel(data = diagnosis_positions, 
                  aes(label = `Final diagnosis`, x = tsne1, y = tsne2),
                  size = 3, 
                  max.overlaps = Inf,  # Show all labels
                  box.padding = 0.5, 
                  point.padding = 0.5, 
                  segment.color = 'grey50') +
  scale_color_manual(values = c("0" = "blue", "1" = "red")) +
  theme_minimal() +
  labs(title = "2D t-SNE Visualization with Euclidean Classifier",
       x = "t-SNE 1",
       y = "t-SNE 2",
       color = "Grok Final") +
  theme(legend.position = "none")
```

# tsne2

```{r}
# Load required libraries
library(dplyr)
library(ggplot2)
library(ggrepel)

# Assuming atlasData contains columns: tsne1, tsne2, `t-SNE classifier`, grok_final

# Convert NA values to a specific category or handle them appropriately
atlasData$`t-SNE classifier imputed` <- ifelse(is.na(atlasData$`t-SNE classifier`), 
                                               "Unknown", 
                                               as.character(atlasData$`t-SNE classifier`))

# Calculate median positions for each diagnosis for labeling
diagnosis_positions <- atlasData %>%
  group_by(`Final diagnosis`) %>%
  summarise(
    tsne1 = median(tsne1),
    tsne2 = median(tsne2)
  )

# Create the 2D plot with t-SNE classifier
ggplot(atlasData, aes(x = tsne1, y = tsne2)) +
  geom_point(aes(color = `t-SNE classifier imputed`), alpha = 0.7, size = 2) +
  geom_text_repel(data = diagnosis_positions, 
                  aes(label = `Final diagnosis`, x = tsne1, y = tsne2),
                  size = 3, 
                  max.overlaps = Inf,  # Show all labels
                  box.padding = 0.5, 
                  point.padding = 0.5, 
                  segment.color = 'grey50') +
  theme_minimal() +
  labs(title = "2D t-SNE Visualization with t-SNE Classifier",
       x = "t-SNE 1",
       y = "t-SNE 2",
       color = "t-SNE Classifier") +
  scale_color_discrete()  # Changed to discrete scale
```

#random forest

```{r}
# Load required libraries
library(dplyr)
library(ggplot2)
library(ggrepel)

# Assuming atlasData contains columns: tsne1, tsne2, `Random Forest classifier`, grok_final

# Convert NA values to a specific category or handle them appropriately
atlasData$`Random Forest classifier imputed` <- ifelse(is.na(atlasData$`Random Forest classifier`), 
                                                      "Unknown", 
                                                      as.character(atlasData$`Random Forest classifier`))

# Calculate median positions for each diagnosis for labeling
diagnosis_positions <- atlasData %>%
  group_by(`Final diagnosis`) %>%
  summarise(
    tsne1 = median(tsne1),
    tsne2 = median(tsne2)
  )

# Create the 2D plot with Random Forest classifier
ggplot(atlasData, aes(x = tsne1, y = tsne2)) +
  geom_point(aes(color = `Random Forest classifier imputed`), alpha = 0.7, size = 2) +
  geom_text_repel(data = diagnosis_positions, 
                  aes(label = `Final diagnosis`, x = tsne1, y = tsne2),
                  size = 3, 
                  max.overlaps = Inf,  # Show all labels
                  box.padding = 0.5, 
                  point.padding = 0.5, 
                  segment.color = 'grey50') +
  theme_minimal() +
  labs(title = "2D t-SNE Visualization with Random Forest Classifier",
       x = "t-SNE 1",
       y = "t-SNE 2",
       color = "Random Forest Classifier") +
  scale_color_discrete()  # Changed to discrete scale
```

