


```{r}
# jan 21 2025
# understand gene annotations

tellMe <- names(mapped_genes) %>% 
  unique()

sample(tellMe, 50)


glimpse(atlasDataClean)
glimpse(geneExpressionData[1:25, 1:5])
glimpse(gene_annotations)
rownames(geneExpressionData[1:25, 1:5])

# jan 23 2025

# Load necessary library
library(purrr)

# Assuming 'mapped_genes' is a list or a vector where each element is like mapped_genes[1]
result <- map(mapped_genes, function(x) {
  if(is.list(x) && length(x) == 1) {
    # If x is a list with one element, apply the operation
    return(length(unlist(x[[1]])))
  } else {
    # If x does not match expected structure, return NA or another fallback value
    return(NA)
  }
})

# Convert the result to integers, assuming all results are numeric or can be converted
integer_values <- as.integer(unlist(result))

# Check if we have 6394 elements
if(length(integer_values) == 6394) {
  print("Successfully processed 6394 elements.")
} else {
  warning("The number of processed elements does not match 6394.")
}

# Print or use integer_values as needed
print(integer_values)

# Calculate the sum of the lengths of all elements in mapped_genes
total_elements <- sum(lengths(mapped_genes))

print(total_elements)

# n = 406318.. ENTREZIDs in multiple categories?
# Combine all elements into one vector, then find unique values and count them
unique_elements <- unique(unlist(mapped_genes))
total_unique_elements <- length(unique_elements)

print(total_unique_elements)

# Combine all elements into one vector
all_values <- unlist(mapped_genes)

# If you want to see the first few elements of this vector for verification:
print(head(all_values))
li(all_values, rownames(geneExpressionData))

# check gene annotations

library(rentrez)

# Function to check if an Entrez ID is valid
check_entrez_id <- function(id) {
  tryCatch({
    # Query NCBI for the gene ID
    result <- entrez_summary(db="gene", id=id)
    return(TRUE)
  }, error = function(e) {
    # If an error occurs (e.g., ID not found), return FALSE
    return(FALSE)
  })
}

# Example IDs, replace with your full list
ids_to_check <- filter(gene_annotations, is.na(SYMBOL)) %>% pull(ENTREZID)

# Check validity
valid_ids <- sapply(ids_to_check, check_entrez_id)
print(valid_ids)

filter(gene_annotations, is.na(SYMBOL)) %>% glimpse()

#####

library(rentrez)

fetch_gene_info <- function(id) {
  result <- entrez_summary(db="gene", id=id)
  return(data.frame(ENTREZID = id, 
                    SYMBOL = result$name, 
                    GENENAME = result$description))
}

# Fetch info for missing IDs
missing_ids <- gene_annotations$ENTREZID[is.na(gene_annotations$SYMBOL)]
new_annotations <- do.call(rbind, lapply(missing_ids, fetch_gene_info))

# Merge with existing annotations
gene_annotations <- rbind(gene_annotations[!is.na(gene_annotations$SYMBOL), ], new_annotations)


##

# library(AnnotationHub)
# Warning in install.packages :
#   package ‘AnnotationHub’ is not available for this version of R
# ah <- AnnotationHub()
# # Query for human gene databases
# query(ah, "Homo sapiens")  # Look for relevant databases
```


# gene groupings

Summary
To find gene groupings:

Use predefined databases (MSigDB, GO, KEGG, Reactome).
Analyze your data for statistical groupings (e.g., variability, DEGs).
Map genes to functional categories with Bioconductor tools.
Create custom groupings based on literature or expert input.
Organize these groupings into a structure suitable for Shiny selectors (lists or data frames).


```{r}


# BiocManager::install("org.Hs.eg.db")

library(org.Hs.eg.db)

# Map Entrez IDs to Gene Symbols and Names
gene_annotations2 <- AnnotationDbi::select(org.Hs.eg.db,
                                          keys = rownames(geneExpressionData),
                                          columns = c("SYMBOL", "GENENAME"),
                                          keytype = "ENTREZID")

head(gene_annotations)


# handle unmapped IDs (none! found)
# unmapped_ids <- setdiff(rownames(geneExpressionData), gene_annotations$ENTREZID)
# print(unmapped_ids)


# Group genes

# biological pathways
# BiocManager::install("clusterProfiler")
library(clusterProfiler)

# Map genes to KEGG pathways (400 error)
# kegg_annotations <- enrichKEGG(gene = rownames(geneExpressionData),
#                                 organism = "hsa", # homo sapiens
#                                 keyType = "ENTREZID")
# head(kegg_annotations)

# gene ontology terms
go_annotations2 <- enrichGO(gene = rownames(geneExpressionData),
                            OrgDb = org.Hs.eg.db,
                            ont = "BP", # Biological Process
                            keyType = "ENTREZID")
head(go_annotations)

# Extract gene-to-GO mapping
# This creates a named list where:
# 
# The names are GO terms (descriptions).
# The values are comma-separated gene IDs.
go_to_genes <- setNames(go_annotations@result$geneID, go_annotations@result$Description)

# Split the gene IDs into vectors
go_to_genes_list <- lapply(go_to_genes, function(x) unlist(strsplit(x, "/")))

# Find overlap with your gene expression data
mapped_genes <- lapply(go_to_genes_list, function(genes) {
  intersect(genes, rownames(geneExpressionData))
})

# Summarize expression by GO terms
go_expression <- sapply(names(mapped_genes), function(go_term) {
  genes <- mapped_genes[[go_term]]
  if (length(genes) > 0) {
    rowMeans(geneExpressionData[genes, , drop = FALSE])  # Mean expression for the term
  } else {
    NA  # No mapped genes for this GO term
  }
})

save(go_expression, gene_annotations, go_annotations, go_to_genes_list, go_to_genes, mapped_genes, file = 'dat/annotations.RData')

```

```{r}
library(pheatmap)

# Remove GO terms with no mapped genes
# go_expression_clean <- go_expression[!is.na(go_expression), ] # threw error
go_expression_clean <- go_expression[!is.na(go_expression)]
str(go_expression_clean) # list. needs to be numeric matrix or df
# Convert the list to a numeric matrix
go_expression_clean <- do.call(rbind, lapply(go_expression_clean, as.numeric))

# Check the resulting structure
str(go_expression_clean)


# Plot heatmap
pheatmap(go_expression_clean, cluster_rows = TRUE, cluster_cols = TRUE,
         main = "GO Term-Level Expression")


library(heatmaply) # way too slow; dont run again

heatmaply(go_expression_clean, 
          k_row = 10,  # Cluster rows into 10 groups
          k_col = 10,  # Cluster columns into 10 groups
          main = "GO Term-Level Expression")

library(ggplot2)

# Summarize mean expression for GO terms
mean_expression <- rowMeans(go_expression_clean)

# Barplot for top 20 GO terms
top_mean_expression <- sort(mean_expression, decreasing = TRUE)[1:20]
ggplot(data = data.frame(GO = names(top_mean_expression), Expression = top_mean_expression),
       aes(x = reorder(GO, Expression), y = Expression)) +
  geom_bar(stat = "identity") +
  coord_flip() +
  labs(title = "Top 20 GO Terms by Mean Expression", x = "GO Term", y = "Mean Expression")

```


```{r}
library(org.Hs.eg.db)
library(clusterProfiler)
library(biomaRt)
library(ggplot2)
library(pheatmap)

# Assuming 'geneExpressionData' is your data frame or matrix with Entrez IDs as rownames

# Fetch gene symbols from Ensembl
ensembl = useMart("ensembl", dataset="hsapiens_gene_ensembl")
geneSymbols <- getBM(attributes = c("entrezgene_id", "hgnc_symbol"),
                     filters = "entrezgene_id", 
                     values = rownames(geneExpressionData), 
                     mart = ensembl)

# Create unique gene symbols to avoid duplicate row name issues
# Here, we ensure uniqueness by considering both Entrez ID and the original symbol
unique_symbols <- paste(geneSymbols$hgnc_symbol, geneSymbols$entrezgene_id, sep = "_")
geneSymbols$unique_symbol <- make.unique(unique_symbols, sep = "_")

# Set row names with these unique symbols
rownames(geneExpressionData) <- setNames(geneSymbols$unique_symbol, 
                                         geneSymbols$entrezgene_id)[rownames(geneExpressionData)]

# Use original Entrez IDs for GO and KEGG analysis
entrez_ids <- rownames(geneExpressionData)

# Perform GO enrichment
go_enrich <- enrichGO(gene = entrez_ids,
                      OrgDb = org.Hs.eg.db,
                      ont = "BP",
                      pAdjustMethod = "BH",
                      pvalueCutoff = 0.05,
                      qvalueCutoff = 0.2,
                      readable = TRUE)

# Display GO enrichment results
print(head(go_enrich, n = 10))

# KEGG pathway analysis
kegg_enrich <- enrichKEGG(gene = entrez_ids,
                          organism = 'hsa',  
                          pvalueCutoff = 0.05)

# Display KEGG results
print(head(kegg_enrich))

# Custom grouping for ZNF genes
znf_genes <- grep("^ZNF", rownames(geneExpressionData), value = TRUE)
if(length(znf_genes) > 0) {
  znf_group <- data.frame(Gene = znf_genes, Group = "ZNF")
  print(znf_group)

  # Visualization of GO enrichment
  print(dotplot(go_enrich, showCategory = 30) + ggtitle("GO Enrichment Analysis"))

  # Heatmap for ZNF genes
  pheatmap(geneExpressionData[znf_genes,], 
           show_rownames = TRUE, 
           main = "Expression of ZNF Genes",
           fontsize_row = 6)
} else {
  print("No ZNF genes found or they might have unique suffixes.")
}
```


```{r}
# cross reference with known dB
# Select a small subset of IDs
test_ids <- head(rownames(geneExpressionData))

# Map these IDs to gene symbols
mapped_genes <- AnnotationDbi::select(org.Hs.eg.db, 
                                      # keys = test_ids, 
                                      keys = rownames(geneExpressionData),
                                      columns = c("SYMBOL", "GENENAME", "GO", "PATH"), 
                                      keytype = "ENTREZID")
print(mapped_genes)

# map genes to GO biological processes

# enrichGO(gene = test_ids, OrgDb = org.Hs.eg.db, ont = "BP")


# merge
# Example of merging annotations with gene expression data
# merged_data <- merge(geneExpressionData, annotations, 
#                      by.x = "rownames", by.y = "ENTREZID", all.x = TRUE)

```


```{r}
library(KEGGREST)

# Convert Entrez IDs to KEGG IDs
entrez_ids <- rownames(geneExpressionData)
kegg_ids <- keggConv("kegg", "hsa")  # Converts human Entrez IDs to KEGG IDs

# Filter your Entrez IDs to match KEGG
valid_kegg_ids <- intersect(entrez_ids, names(kegg_ids))

# Run enrichment analysis with valid KEGG IDs
kegg_annotations <- enrichKEGG(gene = valid_kegg_ids,
                                organism = "hsa", 
                                keyType = "kegg")

```

# examine


```{r cross-reference and explore}

examine <- geneExpressionData %>% 
  select(AGZ_001_U133_2.CEL)

examineOther <- atlasData %>% 
  filter(Filename %in% 'AGZ_001_U133_2.CEL')

rownames(examine) %>% length() %>% unique()

n_distinct(atlasData$Filename) # matches 7375

atlasData %>% 
  group_by(
    Filename
  ) %>% 
  count()

# List of non-tumor categories from the image
non_tumor_categories <- toupper(c("Cerebellum", "Cerebellum, Fetal", "Choroid Plexus", 
                          "CNS", "Nerve Ganglia", "Peri. Nerve", "Pituitary", 
                          "Retina", "Retina, Fetal"))


# Create the binary indicator variable
atlasData2$is_tumor <- ifelse(
  # atlasData2$`Final diagnosis` %in% non_tumor_categories, 
  atlasData2$`Diagnosis class` %in% 'Non-tumor',
  0, 
  1
)

# Check the frequency to ensure the counts match
freq(atlasData2$is_tumor)

sample(atlasData$'Sample ID', 250) %>% freq

atlasDataClean %>% 
  mutate(
    x = str_sub(filename, 1, 3),
    .keep = 'used'
  ) %>% 
  group_by(x) %>% 
  count(sort = T)


```



```{r}

glimpse(geneExpressionData)
glimpse(atlasData)

```




```{r freqs}


map(atlasData[, "Diagnosis class", drop = F], ~ freq(.x))
map(atlasData[, "Diagnosis", drop = F], ~ freq(.x))
map(atlasData[, "Final diagnosis", drop = F], ~ freq(.x))
map(atlasData[, "Age (grouped)", drop = F], ~ freq(.x))

map(select(atlasData, contains('classifier')), drop = F, ~ freq(.x))

map(atlasData[, , drop = F], ~ freq(.x) %>% head())
map(atlasDataClean[, , drop = F], ~ freq(.x) %>% head())
sink('dat/all-freqs'); map(atlasData, ~ freq(.x)); sink()
sink('dat/all-freqs2'); map(atlasDataClean, ~ freq(.x)); sink()
sink('out/all-freqs3'); map(atlasDataClean, ~ freq(.x)); sink()

atlasData %>% filter(as.logical(is_tumor)) %>% pull('Final diagnosis') %>% freq
atlasData %>% filter(as.logical(is_tumor)) %>% pull('Diagnosis class') %>% freq

atlasDataClean %>% 
  pull(diagnosisFinal) %>% 
  freq()
atlasDataClean %>% 
  pull(diagnosis) %>% 
  freq()
atlasDataClean %>% 
  pull(diagnosisClass) %>% 
  freq()


```




# extra



```{r na-check}
library(tidyverse)

# Determine which columns are of double type
double_columns <- geneExpressionData %>%
  summarise(across(everything(), is.double)) %>%
  pivot_longer(everything(), names_to = "column", values_to = "is_double") %>%
  filter(is_double) %>%
  pull(column)

# Check for NA values in these double columns
na_check <- geneExpressionData %>%
  select(all_of(double_columns)) %>%
  summarise(across(everything(), ~any(is.na(.)))) %>%
  pivot_longer(everything(), names_to = "column", values_to = "has_na") %>%
  filter(has_na)

# Display the result
if (nrow(na_check) > 0) {
  print("Columns with NA values in double type:")
  print(na_check$column)
} else {
  print("No NA values found in any double type columns.")
}



```

# henry-adapted


```{r adapt-from-hn}
# #atlas<-readRDS("C:/Users/henry/Downloads/ExpData.rds")
# mean.frame <- data.frame(colMeans(geneExpressionData))
# jde.mean<- data.frame(mean.frame[1:170,])
# # jde.hist <- read.csv("C:/Users/henry/Downloads/hist.csv")
# hist <- read_csv("hist.csv")
# jde.data<- bind_cols(jde.mean,hist)
# colnames(jde.data)[colnames(jde.data) == "mean.frame.1.170..."] <- "Mean"
# boxplot(Mean ~ Histology, data = jde.data,
#         main = "Boxplot by Histology", 
#         xlab = "Histology", 
#         ylab = "mRNA Expression (log2)")
```



```{r adapt-from-hn}


# # jde.surv<- read.csv("C:/Users/henry/Downloads/jde survival.csv")
# jde_survival <- read_csv("jdesurv.csv")
# jde.survival$Age.Groups <- jde_survival$`Age (grouped)`
# 
# jde.survival<- jde_survival[, c("Vital status","Overall survival (months)","Original histology"), drop=F]
# # jde.survival$`Overall survival (months)` <- as.numeric(jde.survival$`Overall survival (months)`)
# fit<- survfit(Surv(`Overall survival (months)`,`Vital status`)~strata(Age.Groups), data=jde.survival)
# plot(fit, xlab="Time", ylab="Survival", main="JDE Data", col=c(1:4), lty=c(1),conf.int = F)
# legend("bottomleft",title = "Age Groups", 0.7, legend=c("10-20","20-40","40-60","60-80"), lty=1, col=1:4)


```





# fig1

```{r}
library(ggplot2)
library(dplyr)

# First, let's summarize the data to get counts for each diagnosis
diagnosis_counts <- atlasData %>%
  count(`Final diagnosis`, sort = TRUE) %>%
  filter(n > 50)  # Only show labels for diagnoses with more than 50 occurrences

# Now, compute the position for labels to avoid overlap
diagnosis_positions <- atlasData %>%
  group_by(`Final diagnosis`) %>%
  summarise(
    tsne1 = median(tsne1),
    tsne2 = median(tsne2)
  ) %>%
  inner_join(diagnosis_counts, by = "Final diagnosis")

# Create the plot
ggplot(atlasData, aes(x = tsne1, y = tsne2)) +
  geom_point(aes(color = `Final diagnosis`), alpha = 0.7, size = 2) +
  geom_text(data = diagnosis_positions, 
            aes(label = `Final diagnosis`, x = tsne1, y = tsne2),
            size = 3, check_overlap = TRUE, vjust = 1, hjust = 1) +  # Adjust size and position
  theme_minimal() +
  labs(title = "t-SNE Visualization of Tumor Types with Diagnosis Labels",
       x = "t-SNE 1",
       y = "t-SNE 2",
       color = "Diagnosis") +
  theme(legend.position = "none")  # Remove legend since we're using labels

# Save the plot if you want
# ggsave("tsne_plot_with_labels.png", width = 15, height = 10, dpi = 300)
```

```{r}
library(ggplot2)
library(dplyr)
library(ggrepel)  # For better label placement

# Compute the position for labels
diagnosis_positions <- atlasData %>%
  group_by(`Final diagnosis`) %>%
  summarise(
    tsne1 = median(tsne1),
    tsne2 = median(tsne2)
  )

# Create the plot with all labels
ggplot(atlasData, aes(x = tsne1, y = tsne2)) +
  geom_point(aes(color = `Final diagnosis`), alpha = 0.7, size = 2) +
  geom_text_repel(data = diagnosis_positions, 
                  aes(label = `Final diagnosis`, x = tsne1, y = tsne2),
                  size = 3, 
                  max.overlaps = Inf,  # Show all labels
                  box.padding = 0.5, 
                  point.padding = 0.5, 
                  segment.color = 'grey50') +
  theme_minimal() +
  labs(title = "t-SNE Visualization of Tumor Types with All Diagnosis Labels",
       x = "t-SNE 1",
       y = "t-SNE 2") +
  theme(legend.position = "none")  # Remove legend since we're using labels

# Save the plot if you want
# ggsave("tsne_plot_all_labels.png", width = 15, height = 10, dpi = 300)
```

```{r}
# blue/red no/yes tumor?

# Calculate median positions for labels
diagnosis_positions <- atlasData2 %>%
  group_by(`Final diagnosis`) %>%
  summarise(
    tsne1 = median(tsne1),
    tsne2 = median(tsne2)
  )

# Create the plot with all labels and color by grok_final
ggplot(atlasData2, aes(x = tsne1, y = tsne2)) +
  geom_jitter(aes(color = factor(is_tumor)), alpha = 0.7, size = 2) +
  geom_text_repel(data = diagnosis_positions, 
                  aes(label = `Final diagnosis`, x = tsne1, y = tsne2),
                  size = 3, 
                  max.overlaps = Inf,  # Show all labels
                  box.padding = 0.5, 
                  point.padding = 0.5, 
                  segment.color = 'grey50') +
  scale_color_manual(values = c("0" = "blue", "1" = "red")) +
  theme_minimal() +
  labs(title = "t-SNE Visualization of Tumor Types with All Diagnosis Labels",
       x = "t-SNE 1",
       y = "t-SNE 2",
       color = "Grok Final") +
  theme(legend.position = "none")

```

# euclidean
```{r}
# Load required libraries
library(dplyr)
library(ggplot2)
library(ggrepel)

# Assuming atlasData contains columns: tsne1, tsne2, `Euclidean classifier`, grok_final

# Median imputation for NA values in `Euclidean classifier`
atlasData$`Euclidean classifier imputed` <- ifelse(is.na(atlasData$`Euclidean classifier`), 
                                                   median(atlasData$`Euclidean classifier`, na.rm = TRUE), 
                                                   atlasData$`Euclidean classifier`)

# Calculate median positions for each diagnosis for labeling
diagnosis_positions <- atlasData %>%
  group_by(`Final diagnosis`) %>%
  summarise(
    tsne1 = median(tsne1),
    tsne2 = median(tsne2)
  )

# Create the 2D plot with Euclidean classifier
ggplot(atlasData, aes(x = tsne1, y = tsne2)) +
  geom_point(aes(color = factor(grok_final)), alpha = 0.7, size = 2) +
  geom_text_repel(data = diagnosis_positions, 
                  aes(label = `Final diagnosis`, x = tsne1, y = tsne2),
                  size = 3, 
                  max.overlaps = Inf,  # Show all labels
                  box.padding = 0.5, 
                  point.padding = 0.5, 
                  segment.color = 'grey50') +
  scale_color_manual(values = c("0" = "blue", "1" = "red")) +
  theme_minimal() +
  labs(title = "2D t-SNE Visualization with Euclidean Classifier",
       x = "t-SNE 1",
       y = "t-SNE 2",
       color = "Grok Final") +
  theme(legend.position = "none")
```

# tsne2

```{r}
# Load required libraries
library(dplyr)
library(ggplot2)
library(ggrepel)

# Assuming atlasData contains columns: tsne1, tsne2, `t-SNE classifier`, grok_final

# Convert NA values to a specific category or handle them appropriately
atlasData$`t-SNE classifier imputed` <- ifelse(is.na(atlasData$`t-SNE classifier`), 
                                               "Unknown", 
                                               as.character(atlasData$`t-SNE classifier`))

# Calculate median positions for each diagnosis for labeling
diagnosis_positions <- atlasData %>%
  group_by(`Final diagnosis`) %>%
  summarise(
    tsne1 = median(tsne1),
    tsne2 = median(tsne2)
  )

# Create the 2D plot with t-SNE classifier
ggplot(atlasData, aes(x = tsne1, y = tsne2)) +
  geom_point(aes(color = `t-SNE classifier imputed`), alpha = 0.7, size = 2) +
  geom_text_repel(data = diagnosis_positions, 
                  aes(label = `Final diagnosis`, x = tsne1, y = tsne2),
                  size = 3, 
                  max.overlaps = Inf,  # Show all labels
                  box.padding = 0.5, 
                  point.padding = 0.5, 
                  segment.color = 'grey50') +
  theme_minimal() +
  labs(title = "2D t-SNE Visualization with t-SNE Classifier",
       x = "t-SNE 1",
       y = "t-SNE 2",
       color = "t-SNE Classifier") +
  scale_color_discrete()  # Changed to discrete scale
```

#random forest

```{r}
# Load required libraries
library(dplyr)
library(ggplot2)
library(ggrepel)

# Assuming atlasData contains columns: tsne1, tsne2, `Random Forest classifier`, grok_final

# Convert NA values to a specific category or handle them appropriately
atlasData$`Random Forest classifier imputed` <- ifelse(is.na(atlasData$`Random Forest classifier`), 
                                                      "Unknown", 
                                                      as.character(atlasData$`Random Forest classifier`))

# Calculate median positions for each diagnosis for labeling
diagnosis_positions <- atlasData %>%
  group_by(`Final diagnosis`) %>%
  summarise(
    tsne1 = median(tsne1),
    tsne2 = median(tsne2)
  )

# Create the 2D plot with Random Forest classifier
ggplot(atlasData, aes(x = tsne1, y = tsne2)) +
  geom_point(aes(color = `Random Forest classifier imputed`), alpha = 0.7, size = 2) +
  geom_text_repel(data = diagnosis_positions, 
                  aes(label = `Final diagnosis`, x = tsne1, y = tsne2),
                  size = 3, 
                  max.overlaps = Inf,  # Show all labels
                  box.padding = 0.5, 
                  point.padding = 0.5, 
                  segment.color = 'grey50') +
  theme_minimal() +
  labs(title = "2D t-SNE Visualization with Random Forest Classifier",
       x = "t-SNE 1",
       y = "t-SNE 2",
       color = "Random Forest Classifier") +
  scale_color_discrete()  # Changed to discrete scale
```

